### 2022-09-04 main/chapter_searches.tex
add draft p-value eqn
more boring text on p-values

% test statistics
A statistic is a function of data, and a test statistic is just a statistic
that is used in statistical tests.
For \atlas\ collision data, event variables are statistics; when working with
data, one always uses statistics and is free to design them for their benefits
in context.
Just as we define event variables (like $\pt$) to remove nuisance
dependencies (on $\phi$ rotations or longitudinal boosts),
useful test statistics can also eliminate unwanted parameters.
The quintessential example is $t$-statistic
\begin{equation}
t = (\bar{\mu} - \mu)\sqrt{\frac{n}{\bar{\sigma}^2}}
,
\end{equation}
where $\mu$ is the known data mean, and $\bar{\mu}$ and $\bar{\sigma}^2$ are
standard estimates of the mean and variance of the data sample respectively.
By dividing out a standard deviation, $t$ is independent of any common
scale factors on the data~\cite{student1908, fisher1925t, lehmann2011fisher}.

\begin{figure}[tp]
\centering
\includegraphics[width=0.46\textwidth]{figures/searches_significant_shrink.png}
\\
\begin{footnotesize}
`So, uh, we did the green study again and got no link. It was probably a{-}{-}'
\\
`RESEARCH CONFLICTED ON GREEN JELLY BEAN ACNE LINK; MORE~STUDY~RECOMMENDED!'
\end{footnotesize}
\caption[
``Significant'' by Randall~Munroe
]{%
``Significant'' by Randall~Munroe~\cite{xkcd2011significant}.
}
\label{fig:searches_significant}
\end{figure}

% p-values
\TODO[
(probable error?), known in modern jargon as $p$-values
tail area = significance
~\cite{fisher1925smrw}
}

\TODO{maximum likelihood}
\TODO{p-value}
\TODO{p-value:
The chance of a small $p$-value is small.
This is true, but not interesting because the value of any $p$-value is small.
By construction $p$-values are uniformly distributed on the unit interval
under a given hypothesis.
The frequency of $p < 0.05$ is $5\%$, but so also is the frequency of
$0.6 < p < 0.65$.
The frequency of $p < \varepsilon$ is identical to the frequency
of $0.6 < p < 0.6 + \varepsilon$.
An appropriate choice of test statistic is therefore necessary for the
$p$-value to have meaning.
}
Figure~\ref{fig:searches_significant}

@@

``%
There is something horrifying in the ideological movement represented by the
doctrine that reasoning, properly speaking, cannot be applied to empirical data
to lead to inferences valid in the real world.%
''~\cite{fisher1956statistical}

\paragraph{Neyman-Pearson.}
\TODO{
Nayman-Pearson:
Seek good test statistics --
likelihood ratio intuitive for two-hypothesis comparison
prove it optimal in a reasonable sense
idea of long-run mostly not wrong
recede to maximum likelihood ratios
~\cite{neymanpearson1928max}
``for which hypothesis A is more probable and those for which it is less
probable (sic!)''
unbiased (unbiassed) test differs from unbiasedness of estimator
``Then the two corresponding densities at the sample point
are (Al) 6-42 x 10-20, (A2) 5 18*x 1O-4 and the ratio of the likelihood of Hypothesis
Al to- that of A2 is 1P24 x 10-16. This ratio confirms the common-sense judgment
that A2 is far more plausible than A1''
\^ Another posterior argument --- this likelihood ratio in in fact a change
in plausibilities. For example, if A2 is an inconsistent, self-contradictory
theory, it is totally implausible, and not more plausible despite predicting
these data better.
``The statistician can
balance the numerical verdict of likelihood against the vaguer expectations de-
rived from `a priori' considerations, and it must be left to his judgment to
at what point the evidence in favour of alternative hypotheses becomes so con-
vincing that Hypothesis A must be rejected.''
Introduce maximum likelihood ratio as interesting without claiming it to be
necessarily best.
Neyman-Pearson lemma --- reproduce Fig 5?
~\cite{neymanpearson1933lemma}
}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.6\textwidth]{figures/searches_atlas_higgs_4l_2207_00320.pdf}
\caption[
Signal and background distributions reproduced from the recent \atlas\
measurement of Higgs boson decays to $4\ell$
]{%
Signal and background distributions reproduced from the recent \atlas\
measurement of Higgs boson decays to $4\ell$~\cite{ATLAS:2022net}.
}
\label{fig:atlas_higgs_4l}
\end{figure}

Figure~\ref{fig:atlas_higgs_4l}

>>>>

\TODO{estimators briefly here}
\TODO{maximum likelihood}

@@

% test statistics
A statistic is a function of data, and a test statistic is just a statistic
that is used in statistical tests.
For \atlas\ collision data, event variables are statistics; when working with
data, one always uses statistics and is free to design them for their benefits
in context.
Just as we define event variables (like $\pt$) to remove nuisance
dependencies (on $\phi$ rotations or longitudinal boosts),
useful test statistics can also eliminate unwanted parameters.
The quintessential example is $t$-statistic
\begin{equation}
t(d) = \frac{\bar{\mu}(d) - \mu}{\bar{\sigma}(d)}\sqrt{n}
,
\end{equation}
where $\mu$ is the known data mean, and $\bar{\mu}(d)$ and $\bar{\sigma}(d)$
are standard estimates of the mean and standard deviation of the data sample
respectively~\cite{student1908, fisher1925t}.
By dividing out a (scale equivariant) standard deviation, $t$ is independent
of common scale factors on the data, and that scale invariance has helped to
drive the study and common use of this $t$ statistic~\cite{lehmann2011fisher}.

One can usefully reduce observed data to just the test statistic
$\tilde{t} = t(d)$ and compare likelihoods to compare how well different
hypotheses predict the observed $\tilde{t}$.
For reasons I cannot explain, Fisher advocates a different use ---
rather than considering likelihoods and likelihood ratios, he recommends
that researchers consider the ``probable error''~\cite{fisher1921probable},
which is a cumulative distribution function in the negated test statistic.
In modern jargon, a `probable error' at the observed test statistic
$\tilde{t}$ and an hypothesis $h$ is a \textbf{$p$-value}
\begin{align}
P(h)
&= \prob{t \geq \tilde{t}}{h} \\
&= L(h) + \prob{t > \tilde{t}}{h}
\label{eqn:searches_pvalue_likelihood}
;
\end{align}
the $p$-value $P(h)$ is the prior probability of data
\emph{greater or equal to} the observed data when ordered by the value of the
test statistic.
Equation~\ref{eqn:searches_pvalue_likelihood} emphasizes that it is just the
likelihood plus some fluff from a tail of larger test statistics.

\TODO{quote EX.II from ~\cite{fisher1921probable}
}

\TODO{p-value:
The chance of a small $p$-value is small.
This is true, but not interesting because the value of any $p$-value is small.
By construction $p$-values are uniformly distributed on the unit interval
under a given hypothesis.
The frequency of $p < 0.05$ is $5\%$, but so also is the frequency of
$0.6 < p < 0.65$.
The frequency of $p < \varepsilon$ is identical to the frequency
of $0.6 < p < 0.6 + \varepsilon$.
An appropriate choice of test statistic is therefore necessary for the
$p$-value to have meaning.
}

\begin{figure}[tp]
\centering
\includegraphics[width=0.6\textwidth]{figures/searches_atlas_higgs_4l_2207_00320.pdf}
\caption[
Signal and background distributions reproduced from the recent \atlas\
measurement of Higgs boson decays to $4\ell$
]{%
Signal and background distributions reproduced from the recent \atlas\
measurement of Higgs boson decays to $4\ell$~\cite{ATLAS:2022net}, with
observed data.
This example illustrates $p$-values pathological dependency on tail selections,
and the benefits of likelihood ratios.
}
\label{fig:atlas_higgs_4l}
\end{figure}

Fisher claims that the $p$-value indicates the ``significance'' of an
observation~\cite{fisher1925smrw}, with small or large values being
suspiciously significant, and that
``If $P$ is between $.1$ and $.9$ there is certainly no reason to suspect the
hypothesis tested.''~\cite{fisher1925smrw}
This may have been true for his intended applications, but it is not general.

Consider for example the distributions of signal (Higgs boson) and background
(mostly $ZZ^*$) samples in Figure~\ref{fig:atlas_higgs_4l}.
If an event is observed with $m_{4\ell} = 125\,\eV[G]$, is it signal or
background?
It is obviously a Higgs boson candidate, but $m_{4\ell}$ is around the middle
of both signal and background distributions, so with an $m_{4\ell}$ test
statistic $P(\textrm{Higgs boson}) \approx P(ZZ^*) \approx 0.5$, so neither
is ``suspect''.
Local ratios between the two distributions clearly favour signal for this
event, however; the $p$-values' failure is in their dependence on irrelevant
tail areas, which are a feature of the histogram and not the data.
In fact, this $ZZ^*$ density peaks for $ZZ$ above a
$2 m(Z) \approx 180\,\eV[G]$
edge, and $Z \to 4\ell$ around $m_{4\ell} \approx m(Z)$;
expanding the histogram to show either of these tails would change nothing
local to our $m_{4\ell} = 125\,\eV[G]$ event, but would change the $p$-values.

Better test statistics can soften this problem, and a key observation by
Neyman and Pearson is an optimality of likelihood ratio test statistics.

\TODO{
``%
There is something horrifying in the ideological movement represented by the
doctrine that reasoning, properly speaking, cannot be applied to empirical data
to lead to inferences valid in the real world.%
''~\cite{fisher1956statistical}
}

\paragraph{Neyman-Pearson.}
\begin{figure}[tp]
\centering
\includegraphics[width=0.46\textwidth]{figures/searches_significant_shrink.png}
\\
\begin{footnotesize}
`So, uh, we did the green study again and got no link. It was probably a{-}{-}'
\\
`RESEARCH CONFLICTED ON GREEN JELLY BEAN ACNE LINK; MORE~STUDY~RECOMMENDED!'
\end{footnotesize}
\caption[
``Significant'' by Randall~Munroe
]{%
``Significant'' by Randall~Munroe~\cite{xkcd2011significant}.
}
\label{fig:searches_significant}
\end{figure}
Figure~\ref{fig:searches_significant}

\TODO{
Nayman-Pearson:
Seek good test statistics --
likelihood ratio intuitive for two-hypothesis comparison
prove it optimal in a reasonable sense
idea of long-run mostly not wrong
recede to maximum likelihood ratios
~\cite{neymanpearson1928max}
``for which hypothesis A is more probable and those for which it is less
probable (sic!)''
unbiased (unbiassed) test differs from unbiasedness of estimator
``Then the two corresponding densities at the sample point
are (Al) 6-42 x 10-20, (A2) 5 18*x 1O-4 and the ratio of the likelihood of Hypothesis
Al to- that of A2 is 1P24 x 10-16. This ratio confirms the common-sense judgment
that A2 is far more plausible than A1''
\^ Another posterior argument --- this likelihood ratio in in fact a change
in plausibilities. For example, if A2 is an inconsistent, self-contradictory
theory, it is totally implausible, and not more plausible despite predicting
these data better.
``The statistician can
balance the numerical verdict of likelihood against the vaguer expectations de-
rived from `a priori' considerations, and it must be left to his judgment to
at what point the evidence in favour of alternative hypotheses becomes so con-
vincing that Hypothesis A must be rejected.''
Introduce maximum likelihood ratio as interesting without claiming it to be
necessarily best.
Neyman-Pearson lemma --- reproduce Fig 5?
~\cite{neymanpearson1933lemma}
}

@@
